---
title: "World Population Data - Data Prep & EDA"
author: "Fahim Ahamed"
date: "2025-09-25"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'E:/CCNY/DSE I1030 - Applied Statistics/Project 1')
```

```{r}
library(tidyverse)
library(ggplot2)
library(countrycode)
library(patchwork)
library(flextable)
```

#Data Processing & Cleaning

```{r}
wrangle_data <- function(file_path, value_name) {
  read.csv(file_path) %>%
    pivot_longer(
      cols = starts_with('X'),
      names_to = 'Year',
      values_to = value_name,
      values_drop_na = TRUE
    ) %>%
    mutate(Year = as.integer(gsub('X', '', Year))) %>%
    select(
      Country.Name,
      Country.Code,
      Year,
      all_of(value_name)
    )
}

population_df <- wrangle_data("data/pop_data.csv", "Population")
birth_rate_df <- wrangle_data("data/birth_rate.csv", "BirthRate")
death_rate_df <- wrangle_data("data/death_rate.csv", "DeathRate")

```

```{r}
wrangle_aux_data <- function(file_path, value_name) {
  read_csv(file_path) %>%
    pivot_longer(
      cols = -c(`Country Name`, `Country Code`, `Indicator Name`, `Indicator Code`),
      names_to = "Year",
      values_to = value_name,
      values_drop_na = TRUE
    ) %>%
    mutate(Year = as.integer(Year)) %>%
    select(
      Country.Name = `Country Name`, 
      Country.Code = `Country Code`, 
      Year, 
      all_of(value_name)
    )
}

migration_df <- wrangle_aux_data("data/migration_data.csv", "Net_Migration")
working_age_df <- wrangle_aux_data("data/age_data.csv", "Percent_15_64")

country_meta <- read.csv("data/metadata.csv")
valid_country_codes <- countrycode::codelist$iso3c %>% na.omit() %>% unique()
```

```{r}
data_merged <- population_df %>%
  full_join(birth_rate_df, by = c("Country.Name", "Country.Code", "Year")) %>%
  full_join(death_rate_df, by = c("Country.Name", "Country.Code", "Year")) %>%
  full_join(migration_df, by = c("Country.Name", "Country.Code", "Year")) %>%
  full_join(working_age_df, by = c("Country.Name", "Country.Code", "Year")) %>%
  merge(country_meta %>% select(Country.Code, Region, IncomeGroup), by = "Country.Code") %>%
  mutate(Net_Migration = as.numeric(Net_Migration)) %>%
  mutate(Percent_15_64 = as.numeric(Percent_15_64)) %>%
  mutate(IncomeGroup = factor(IncomeGroup,
    levels = c("Low income", "Lower middle income", 
    "Upper middle income", "High income"))) %>%
  filter(if_all(everything(), ~ !is.na(.))) %>%
  filter(IncomeGroup != "" & !is.na(IncomeGroup)) %>%
  filter(Country.Code %in% valid_country_codes) %>%
  arrange(Country.Name, Year)

colnames(data_merged)[colnames(data_merged) == "Country.Code"] <- "Code"
colnames(data_merged)[colnames(data_merged) == "Country.Name"] <- "Name"
```

```{r}
head(data_merged, 3) %>%
  flextable() %>%
  set_caption("World population (first 3 rows of merged datasets)") %>%
  fontsize(size = 6, part = "all")
```


```{r}
data_final <- data_merged %>%
  group_by(Name) %>%
  arrange(Name, Year) %>%
  mutate(Lambda = Population/lag(Population)) %>%
  mutate(R = (BirthRate - DeathRate) / 1000) %>%
  ungroup()

data <- data_final[data_final$Year != 1960, ]

head(data, 3)
```

## Part 1: EDA of population growth.

```{r}
mean_growth <- mean(data$Lambda, na.rm = TRUE)
min_growth <- min(data$Lambda, na.rm = TRUE)
max_growth <- max(data$Lambda, na.rm = TRUE)

yearly_totals <- data %>%
  group_by(Year) %>%
  summarise(Total_Population = sum(Population, na.rm = TRUE))

p1 <- ggplot(yearly_totals, aes(x = Year, y = Total_Population / 1e9)) +
  geom_point(color = 'steelblue', size = 1.5) +
  ggtitle('Total Population') +
  xlab('Year') +
  ylab('Total Population (billions)') +
  theme_minimal()

p2 <- ggplot(data, aes(x = Lambda)) +
  geom_histogram(binwidth = 0.02, fill = 'steelblue', color = 'black') +
  geom_vline(aes(xintercept = mean_growth), color = 'red', linetype = 'dashed', linewidth = 1) +
  scale_x_continuous(limits = c(min_growth, max_growth)) +
  ggtitle('Annual Growth Rates') +
  xlab('Growth Rate') +
  ylab('Frequency')

p1 + p2
```
### Which countries have experienced the highest increase/decrease in population over time?

```{r}
growth_all_years <- data %>%
  group_by(Name) %>%
  summarise(mean_growth = mean(Lambda, na.rm = TRUE))

N_disp <- 9
top_countries_all_years <- growth_all_years %>% arrange(desc(mean_growth)) %>% head(n = N_disp)

top_growth_countries <- top_countries_all_years$Name

data %>%
  filter(Name %in% top_growth_countries) %>%
  ggplot(aes(x = Year, y = Population / 1e6, color = Name)) +
  geom_line(size = 0.8, show.legend = FALSE) +
  facet_wrap(~ Name, scales = 'free_y', ncol = 3) +
  labs(title = 'Countries with highest average increase in population over time',
       x = 'Year', y = 'Population (millions)') +
  theme_minimal()
```
```{r}
tail_countries_all_years <- growth_all_years %>% arrange(mean_growth) %>% head(n = N_disp)
tail_growth_countries <- tail_countries_all_years$Name

data %>%
  filter(Name %in% tail_growth_countries) %>%
  ggplot(aes(x = Year, y = Population / 1e6, color = Name)) +
  geom_line(size = 0.8, show.legend = FALSE) +
  facet_wrap(~ Name, scales = 'free_y', ncol = 3) +
  labs(title = 'Countries with highest average decrease in population over time',
       x = 'Year',
       y = 'Population (millions)') +
  theme_minimal()
```
### Which countries have been experiencing the highest increase/decrease in population in the last five years?

```{r}
growth_last_five_years <- data %>%
  filter(Year >= 2020 & Year <= 2024) %>%
  group_by(Name) %>%
  summarise(
    Pop_2020 = first(Population[order(Year)]),
    Pop_2024 = last(Population[order(Year)]),
    Growth_Percent = (Pop_2024 - Pop_2020) / Pop_2020 * 100
  )

top_countries_last_five_years <- growth_last_five_years %>% arrange(desc(Growth_Percent)) %>% head(n = N_disp)

top_countries_last_five_years %>%
  flextable() %>%
  set_caption('Top Countries highest increase in population in last 5 years') %>%
  autofit()

tail_countries_last_five_years <- growth_last_five_years %>% arrange(Growth_Percent) %>% head(n = N_disp)

tail_countries_last_five_years %>%
  flextable() %>%
  set_caption('Countries with highest decrease in population in last 5 years') %>%
  autofit()
```

### Histograms of annual growth rates with highest average rates.

```{r}
top_countries_avg <- growth_all_years %>% arrange(desc(mean_growth)) %>% head(n = N_disp)

ggplot(data %>% filter(Name %in% top_countries_avg$Name), aes(x = Lambda)) +
  geom_histogram(binwidth = 0.02, fill = 'steelblue', color = 'black') +
  scale_x_continuous(limits = c(min_growth, max_growth)) +
  geom_vline(aes(xintercept = mean_growth), color = 'red', linetype = 'dashed', linewidth = 1) +
  facet_wrap(~ Name, scales = 'free_y', ncol = 3) +
  labs(x = 'Annual Growth Rate', y = 'Frequency')
```

### Histograms of annual growth rates with lowest average rates.

```{r}
tail_countries_avg <- growth_all_years %>% arrange(mean_growth) %>% head(n = N_disp)

ggplot(data %>% filter(Name %in% tail_countries_avg$Name), aes(x = Lambda)) +
  geom_histogram(binwidth = 0.02, fill = 'lightblue', color = 'black') +
  scale_x_continuous(limits = c(min_growth, max_growth)) +
  geom_vline(aes(xintercept = mean_growth), color = 'red', linetype = 'dashed', linewidth = 1) +
  facet_wrap(~ Name, scales = 'free_y', ncol = 3) +
  labs(x = 'Annual Growth Rate', y = 'Frequency')
```

### Population growth of countries over time grouped by region.

```{r fig.width=8, fig.height=4}
regional_population <- data %>%
  group_by(Region, Year) %>%
  summarise(Total_Population = sum(Population, na.rm = TRUE), .groups = 'drop')

ggplot(regional_population, aes(x = Year, y = Total_Population, color = Region)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = 'Total Population Growth by Region',
    x = 'Year',
    y = 'Total Regional Population',
    color = 'Region'
  ) +
  scale_y_continuous(labels = scales::label_number(scale_cut = scales::cut_si(''))) +
  theme_minimal()
```

```{r fig.width=10, fig.height=5}
regional_growth <- data %>%
  group_by(Region, Year) %>%
  summarise(mean_reg_rates = mean(Lambda, na.rm = TRUE), .groups = 'drop')


ggplot(regional_growth, aes(x = Year, y = mean_reg_rates, color = Region)) +
  geom_line(linewidth = 1.2) + 
  labs(
    title = "Average Growth Rate by Region",
    x = "Year", 
    y = "Population Growth Rate",
    color = "Region"
  ) +
  theme_minimal()
```

```{r fig.width=10}
ggplot(data, aes(x = Region, y = Lambda, fill = Region)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(limits = c(0.95, 1.075)) +
  xlab("Region") +
  ylab("Population Growth Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```
### Population growth of countries over time grouped by income group.

```{r fig.width = 8, fig.height = 4}
income_population <- data %>%
  group_by(IncomeGroup, Year) %>%
  summarise(Total_Population = sum(Population, na.rm = TRUE), .groups = 'drop')


ggplot(income_population, aes(x = Year, y = Total_Population, color = IncomeGroup)) +
  geom_line(linewidth = 1.2) + 
  
  labs(
    title = "Total Population Growth by Income Group",
    x = "Year", 
    y = "Total Regional Population",
    color = "IncomeGroup")  +
  theme_minimal()
```
```{r fig.width = 8, fig.height = 4}
inc_gr_growth <- data %>%
  group_by(IncomeGroup, Year) %>%
  summarise(mean_Inc_rates = mean(Lambda, na.rm = TRUE), .groups = 'drop')


ggplot(inc_gr_growth, aes(x = Year, y = mean_Inc_rates, color = IncomeGroup)) +
  geom_line(linewidth = 1.2) + 
  labs(
    title = "Average Growth Rate by Income Group",
    x = "Year", 
    y = "Population Growth Rate",
    color = "IncomeGroup"
  ) +
  theme_minimal()
```
```{r}
ggplot(data, aes(x = IncomeGroup, y = Lambda, fill = IncomeGroup)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(limits = c(0.95, 1.075)) +
  xlab("Income Group") +
  ylab("Population Growth Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## Part 2: Exponential population growth model

### Countries with highest instantaneous rate over time.

```{r}
rates_all_years <- data %>%
  group_by(Name) %>%
  summarise(mean_inst_rates = mean(R, na.rm = TRUE))

mean_R <- median(data$R, na.rm = TRUE)
min_R <- min(data$R, na.rm = TRUE)
max_R <- max(data$R, na.rm = TRUE)

top_countries_all_years <- rates_all_years %>% arrange(desc(mean_inst_rates)) %>% head(n = N_disp)

top_growth_countries <- top_countries_all_years$Name

data %>%
  filter(Name %in% top_growth_countries) %>%
  ggplot(aes(x = Year, y = Population / 1e6, color = Name)) +
  geom_line(size = 0.8, show.legend = FALSE) +
  facet_wrap(~ Name, scales = 'free_y', ncol = 3) +
  labs(title = 'Countries with highest average instantaneous rate over time',
       x = 'Year', y = 'Population (millions)') +
  theme_minimal()
```
### Countries with lowest instantaneous rate over time.

```{r}
tail_countries_all_years <- rates_all_years %>% arrange(mean_inst_rates) %>% head(n = N_disp)

bottom_growth_countries <- tail_countries_all_years$Name

data %>%
  filter(Name %in% bottom_growth_countries) %>%
  ggplot(aes(x = Year, y = Population / 1e6, color = Name)) +
  geom_line(size = 0.8, show.legend = FALSE) +
  facet_wrap(~ Name, scales = 'free_y', ncol = 3) +
  labs(title = 'Countries with lowest average instantaneous rate over time',
       x = 'Year', y = 'Population (millions)') +
  theme_minimal()
```
### Which countries have been experiencing the highest/lowest instantaneous rate in the last five years?

```{r}
rates_last_5 <- data %>%
  filter(Year >= 2020 & Year <= 2024) %>%
  group_by(Name) %>%
  summarise(
    Pop_2020 = first(Population[order(Year)]),
    Pop_2024 = last(Population[order(Year)]),
    mean_inst_rates = mean(R, na.rm = TRUE)
  )

top_countries_last_five_years <- rates_last_5 %>% arrange(desc(mean_inst_rates)) %>% head(n = N_disp)

top_countries_last_five_years %>%
  flextable() %>%
  set_caption('Top Countries highest increase in population in last 5 years') %>%
  autofit()

tail_countries_last_five_years <- rates_last_5 %>% arrange(mean_inst_rates) %>% head(n = N_disp)

tail_countries_last_five_years %>%
  flextable() %>%
  set_caption('Countries with highest decrease in population in last 5 years') %>%
  autofit()
```

### Histograms of instantaneous rates with highest average rates over time.

```{r}
top_countries_avg <- rates_all_years %>% arrange(desc(mean_inst_rates)) %>% head(n = N_disp)

ggplot(data %>% filter(Name %in% top_countries_avg$Name), aes(x = R)) +
  geom_histogram(bins = 20, fill = 'steelblue', color = 'black') +
  scale_x_continuous(limits = c(min_R, max_R)) +
  geom_vline(aes(xintercept = mean_R), color = 'red', linetype = 'dashed', linewidth = 1) +
  facet_wrap(~ Name, scales = 'free_y', ncol = 3) +
  labs(x = 'Annual Instantaneous Rate', y = 'Frequency')
```

### Histograms of instantaneous rates with lowest average rates over time.

```{r}
tail_countries_avg <- rates_all_years %>% arrange(mean_inst_rates) %>% head(n = N_disp)

ggplot(data %>% filter(Name %in% tail_countries_avg$Name), aes(x = R)) +
  geom_histogram(bins = 20, fill = 'lightblue', color = 'black') +
  scale_x_continuous(limits = c(min_R, max_R)) +
  geom_vline(aes(xintercept = mean_R), color = 'red', linetype = 'dashed', linewidth = 1) +
  facet_wrap(~ Name, scales = 'free_y', ncol = 3) +
  labs(x = 'Annual Instantaneous Rate', y = 'Frequency')
```

### Instantaneous rates of countries over time grouped by region.

```{r fig.width=8, fig.height=4}
regional_rates <- data %>%
  group_by(Region, Year) %>%
  summarise(mean_reg_rates = mean(R, na.rm = TRUE), .groups = 'drop')

ggplot(regional_rates, aes(x = Year, y = mean_reg_rates, color = Region)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = 'Average Instantaneous rates per Region',
    x = 'Year',
    y = 'Average Instantaneous Rate',
    color = 'Region'
  ) +
  theme_minimal()
```
```{r fig.width = 8}
ggplot(data, aes(x = Region, y = R, fill = Region)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(limits = c(-0.05, 0.05)) +
  xlab("Region") +
  ylab("Instant Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```
### Instantaneous rates of countries by income group.

```{r fig.width = 8}
inc_gr_rates <- data %>%
  group_by(IncomeGroup, Year) %>%
  summarise(mean_Inc_rates = mean(R, na.rm = TRUE), .groups = 'drop')

ggplot(inc_gr_rates, aes(x = Year, y = mean_Inc_rates, color = IncomeGroup)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = 'Average Instantaneous rates per Income Group',
    x = 'Year',
    y = 'Average Instantaneous Rate',
    color = 'IncomeGroup'
  ) +
  theme_minimal()
```

```{r fig.width = 8, fig.height = 4}
ggplot(data_final, aes(x = IncomeGroup, y = R, fill = IncomeGroup)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(limits = c(-0.05, 0.05)) +
  xlab("Income Group") +
  ylab("Instantaneous Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

## Comparing the results in Parts 1 & 2

```{r fig.width = 8}
N_years <- yearly_totals$Year[length(yearly_totals)] - yearly_totals$Year[1]

t <- yearly_totals$Year
t <- t - t[1] + 1
N_years <- length(t)
P <- yearly_totals$Total_Population
P0 <- P[1]
PL <- P[N_years]
geom_ave_growth <- (PL/P0)^(1/N_years)

yearly_totals <- yearly_totals %>%
  mutate(
    pop_est1 = P0 * mean_growth^t,
    pop_est2 = P0 * geom_ave_growth^t,
    pop_est3 = P0 * (exp(mean_R))^t
    )

plot(yearly_totals$Year, yearly_totals$Total_Population / 1e9, col="steelblue",
     xlab = "Year", ylab= "Total Population (billions)") 
lines(yearly_totals$Year, yearly_totals$pop_est1 / 1e9, 
       col = "blue")
lines(yearly_totals$Year, yearly_totals$pop_est2 / 1e9, 
       col = "darkgreen")
lines(yearly_totals$Year, yearly_totals$pop_est3 / 1e9, 
       col = "salmon")
legend('topleft', legend = c('Original Population', 'Arithmetic Mean Growth', 'Geometric Mean Growth', 'Exponential Growth'),
       col = c('steelblue', 'blue', 'darkgreen', 'salmon'), pch = 19)
grid()
```

```{r fig.width = 8, fig.height = 6}
df_compare <- merge(rates_all_years, growth_all_years, by = "Name")

p2 <- ggplot(df_compare, aes(x = mean_inst_rates, y = log(mean_growth)) ) +
  geom_point(color='blue') +
  labs(title = "Average Instantaneous Rate(r) and ln(Growth Rate λ)",
       x = "Instantaneous Rate (r)",
       y = "ln(Growth Rate λ)") +
  theme_minimal()

ggMarginal(p2, type = "histogram", fill = "skyblue", color = "darkblue")

CR <- cor(df_compare$mean_inst_rates, log(df_compare$mean_growth))
```

## Other Factors influencing growth: Migration and Age Distribution

### Age Distribution factor

A population with a large proportion of young adults will likely have a high birth rate and experience natural increase, even with moderate fertility.

```{r fig.width = 8, fig.height = 4}
working_age_IG <- data %>%
  group_by(IncomeGroup, Year) %>%
  summarise(mean_working_age = mean(Percent_15_64, na.rm = TRUE), .groups = 'drop')

ggplot(working_age_IG, aes(x = Year, y = mean_working_age, color = IncomeGroup)) +
  geom_line(linewidth = 1.2) +
  labs(
    title = 'Population ages 15-64 (% of total) by Income Group',
    x = 'Year',
    y = 'Population ages 15-64 (% of total)',
    color = 'IncomeGroup'
  ) +
  theme_minimal()
```

```{r fig.width = 8}
ggplot(data_final, aes(x = IncomeGroup, y = Percent_15_64, fill = IncomeGroup)) +
  geom_boxplot(alpha = 0.7) +
  xlab("Income Group") +
  ylab("Population ages 15-64 (% of total)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

### Migration factor

In addition to natural increase, population growth is also affected by immigration (people entering a population) and emigration (people leaving a population). 

"Net migration" is the difference between the number of people entering a country or region (immigrants) and the number of people leaving (emigrants) over a specific period. A positive net migration indicates more people are entering than leaving, while a negative number means more people are emigrating. 

```{r fig.width = 8, fig.height = 5}
migration_IG <- data %>%
  group_by(IncomeGroup, Year) %>%
  summarise(mean_migration= mean(Net_Migration, na.rm = TRUE), .groups = 'drop')


ggplot(migration_IG, aes(x = Year, y = mean_migration, color = IncomeGroup)) +
  geom_line(linewidth = 1.2) + 
  labs(
    title = "Net Migration by Income Group",
    x = "Year", 
    y = "Net Migration",
    color = "IncomeGroup"
  ) +
  theme_minimal()
```


```{r fig.width = 8}
ggplot(data_final, aes(x = IncomeGroup, y = Net_Migration, fill = IncomeGroup)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(limits = c(-10^4, 10^4)) +
  xlab("Income Group") +
  ylab("Net Migration") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

The net migration rate expresses this figure as a proportion per 1,000 people, using the formula following formula (https://en.wikipedia.org/wiki/Net_migration_rate): 

Net Migration Rate = 1,000 * (Immigrants - Emigrants) / Mid-Year Population.

```{r}
data <- data %>%
  mutate(
    Migration_Rate = (Net_Migration /Population),
    R_v2 = R + Migration_Rate
    )
```


```{r fig.width = 8, fig.height = 4}
migration_rate_IG <- data %>%
  group_by(IncomeGroup, Year) %>%
  summarise(mean_migration_rate = mean(Migration_Rate, na.rm = TRUE), .groups = 'drop')

ggplot(migration_rate_IG, aes(x = Year, y = mean_migration_rate, color = IncomeGroup)) +
  geom_line(linewidth = 1.2) + 
  labs(
    title = "Net Migration Rate by Income Group",
    x = "Year", 
    y = "Net Migration Rate",
    color = "IncomeGroup"
  ) +
  theme_minimal()
```

```{r fig.width = 8}
ggplot(
  data, aes(x = IncomeGroup, y = Migration_Rate, fill = IncomeGroup)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_continuous(limits = c(-0.02, 0.02)) +
  xlab("Income Group") +
  ylab("Net Migration Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r}
ggMarginal(p2, type = "histogram", fill = "skyblue", color = "darkblue")
```

```{r}
rates_all_years <- data %>%
  group_by(Name) %>%
  summarise(mean_inst_rates = mean(R_v2, na.rm = TRUE))

df_compare <- merge(rates_all_years, growth_all_years, by = "Name")

p3 <- ggplot(df_compare, aes(x = mean_inst_rates, y = log(mean_growth)) ) +
  geom_point(color='blue') +
  labs(title = "Average Modified Instantaneous Rate (r_2) and ln(Growth Rate λ)",
       x = "Instantaneous Rate (r_2)",
       y = "ln(Growth Rate λ)") +
  theme_minimal()

ggMarginal(p3, type = "histogram", fill = "skyblue", color = "darkblue")

CR_v2 <- cor(df_compare$mean_inst_rates, log(df_compare$mean_growth))
```
The correlation coefficient for originally defined Instantaneous growth rate, $r$ was `r CR`. Inclusion of the Net migration in the formula, the correlation coefficient has increased to `r CR_v2`.

```{r}
write.csv(data, "data/cleaned_data.csv")
```
